Composing like a human

### 2019-04-26 paper/main.tex
finished GAN and evaluation metrics

Over the past sixty years, Artificial Neural Networks (ANNs) have experienced a number of cycles in popularity noted by extensive publicity of over-inflated expectations on the promises of connectionism (the field of science that tries to explain mental phenomena with the help of ANNs [CITATION NEEDED]): that a system could learn to do anything it was programmed to do [CITATION NEEDED] \parencite{minsky_perceptrons_1988}. However, each wave of praise was followed by a mass of public disappointment, criticism, and funding cuts [CITATION NEEDED]. These upturns and downturns of the field of connectionism - or more generally A.I. - are now known in the literature as the A.I. winters [CITATION NEEDED].

The recent revival of ANNs under the scope of Deep Learning (DL) has brought new ways to tackle Machine Learning problems, most notably in the context of classification (but also regression, although less popular; see \cite{lecun_deep_2015} for several examples of DL applications). In addition, with the advent of chatbots \parencite{dale_return_2016}, voice-activated personal assistants \parencite{xiong_microsoft_2018}, and general-purpose A.I. systems \parencite{vinyals_starcraft_2017}, DL introduced new frontiers in artificial intelligence research.

% NOTE: USE ME LATER
% Currently, two exciting areas of research in the field of Deep Learning (DL) are in designing generative models \parencite{goodfellow_nips_2016} and training NNs on small datasets \parencite{finn_model-agnostic_2017, nichol_first-order_2018}. This research will combine both of these tasks by exploring the extent to which generative models can create novel and qualitative samples of music under the constraints posed by the few-shot learning problem. In other words, when data is scarce, can a generative model create samples of music that are new (i.e., not in the training set), consistent with common scales, and diverse (tone sequences are not repeated over and over, combines chord structures with single-note sequences, is not restricted to - say - one octave). This research will evaluate the generated samples of two state-of-the-art generative models (C-RNN-GAN of \cite{mogren_c-rnn-gan_2016} and Performance-RNN of \cite{oore_this_2018}) that have been trained under the Reptile learning framework \parencite{nichol_first-order_2018} on small amounts of MIDI data.

Broadly speaking, DL as a field, promises to solve tasks which are easily (intuitively) performed by people, but hard to formalize \parencite[e.g., recognizing faces or spoken word;][]{goodfellow_deep_2016}. The solution comprises of enabling computers to learn form experience and understand the world through the discovery of hierarchical relationships between concepts \parencite{lecun_deep_2015}. This way, there is no need for manual input on the sort of knowledge that the computer needs \parencite{goodfellow_deep_2016}. A more simplistic and concise interpretation of DL would be to see it as a statistical technique for identifying patterns in sample data using large (deep) ANNs \parencite{marcus_deep_2018}.

The reasons for presenting these facts will be clarified in the following sections however, a terse argumentation is that progress towards general A.I. systems is marked by scant explorations of ANNs outside of the classification task, and even more so when trying to overcome some of their shortcomings, such as being highly reliant on massive amounts of data [CITATION NEEDED].
% NOTE: Maybe I was too harsh?

@@

For the purposes of this paper (see Section \ref{sec:thesis_goal}), generative model are central to our experiments (Chapter \ref{chap:methods}). Borrowing the analogy from \textcite{goodfellow_nips_2016}, one might wonder what is the value of studying generative models. After all, in the image domain, such a framework would merely generate more images (something which the Internet as no shortage of). Some of the more eloquent answers include using generative models for simulating possible futures to be later used for planning [CITATION NEEDED], providing predictions to missing data, or enabling work on multi-modal outputs, where a single input may have more than one correct answer \parencite{goodfellow_nips_2016}.

However, successes in generative models have been scarce, with Fully Visible Belief Networks \parencite{frey_does_1996} being the most suitable (and popular) method for the longest run. Yet, this method requires generating each sample one at a time (ramping up the cost of the process to $O(n)$). Coupled with the fact that the computation is done through a neural network, creating $n$ samples requires a substantial amount of time \parencite{goodfellow_nips_2016}. Variational Autoencoders \parencite{kingma_auto-encoding_2013} are another popular method that is used for this sort of task. Their main drawback is that when a weak approximation of a prior or posterior distribution of the data is used, the model might learn something other than the true data distribution \parencite{goodfellow_nips_2016}.

@@

Although not as popular as the abovementioned, Recurrent Neural Networks (RNNs) have been successfully implemented for various generative tasks \parencite{jenal_rnn-based_2019, ICML2011Sutskever_524}. More specifically, a variation of these, Long Short-Term Memory Networks \parencite[LSTMs;][]{hochreiter_long_1997}, are of particular interest to this task. RNNs (and by extension LSTMs) contain a high-dimensional hidden state which allows them to remember and predict from previous inputs \parencite{ICML2011Sutskever_524}. This makes them great candidates for predicting sequences \parencite{graves_generating_2013} of text or characters, where the input at $t_0$ might be influenced by prior input at $t_{-1}$, $t_{-2}$, etc \parencite{fan_tts_2014}.

@@

These developments in generative models lie in the field of Computer Vision, but that is not to say that we can generate only images. GANs (and others) have been applied - with varying degrees of success - in Natural Language Processing where the scope was to generate high-quality text with sufficient diversity \parencite[e.g.,][]{yu_seqgan_2016, chen_adversarial_2018}. Less explored, but still successful, have been the efforts in generating music \parencite[e.g.,][]{mogren_c-rnn-gan_2016, dong_musegan_2017}. This paper will focus on the problem of generating music with a NN-based model. Respectively, this paper will evaluate the samples generated by C-RNN-GAN \parencite{mogren_c-rnn-gan_2016} and Performance-RNN \parencite{oore_this_2018} when trained under the few-shot learning problem (detailed below).

@@

It is fairly well known that DL models tend to be very "data-hungry", a caveat imposed by its optimization algorithm \parencite{chen_closer_2018, ravi_optimization_2016}. DL usually relies on a procedure called stochastic gradient descent (SGC) which implies computing for a given input vector the output and error, followed by calculating the input's average gradient and consequently adjusting the network's weights \parencite{lecun_deep_2015}. \textcite{ravi_optimization_2016} argue that the iterative nature of gradient-optimization algorithms does not allow them to perform well under the constraint of a set number of updates: when faced with non-convex optimization problems, DL algorithms do not guarantee convergence within a reasonable amount of time. Finally, they add by saying that for each new problem (e.g., new dataset), the model needs to reinitialize with a new set of random parameters which hurts the network's ability to converge to a good solution when constrained by a limited number of updates \parencite{ravi_optimization_2016}.

Besides the promise of making DL algorithms easier to train, primary reasons why this issue needs to be addressed are as follows. Firstly, humans have the ability to generalize after one (or few) example(s) of a given object \parencite{vinyals_matching_2016, chen_closer_2018, ravi_optimization_2016}, something which DL algorithms do not. Secondly, this would alleviate the constant need for data collection in order to provide reasonable performances. Finally, there are many fields where the data exhibits a large number of classes, with few examples per class and Bridging the gap between human-type learning and current learning architectures would allow models to properly capture this sort of data \parencite{ravi_optimization_2016, larochelle_few-shot_2017}. This is known as the few-shot problem \parencite{chen_closer_2018, vinyals_matching_2016, larochelle_few-shot_2017}.

\section{Making music like a hooman} \label{sec:thesis_goal}

In spite of its proposed benefits, developments and experiments in few-shot learning are still scarce \parencite{larochelle_few-shot_2017} and evaluations are largely concentrated on image data \parencite[see][]{lake_omniglot_2019, clouatre_figr_2019, vinyals_matching_2016, chen_closer_2018, ravi_optimization_2016}. Few-shot experiments with generative networks are even less frequent \parencite{zhang_metagan_2018, clouatre_figr_2019}. This research aims to explore the viability of this learning approach in the context of generating music data. More concretely, in light of recent developments, this research will evaluate two generative networks modelled on MIDI data under the constraints of few-shot learning, and comparing its performance to the baseline proposed by \textcite{larochelle_few-shot_2017} and properties extracted from real music performances.

\chapter{Related Work}

@@

C-RNN-GAN and Performance-RNN have not been tested and adapted to the few-shot learning problem, but more general GANs and LSTM networks have been \parencite[see][]{zhang_metagan_2018, vinyals_matching_2016}. However, as mentioned previously, no evaluation has been made in the musical domain. Therefore, this study will seek to answer the following research questions:

\begin{itemize}
    \item To what extent is the music created by a few-shot generative model comparable to the music of a generative model that is trained on the entire dataset?
    \item To what extent is the music created by a few-shot generative model comparable to real music?

\end{itemize}

@@

\chapter{Results}

\chapter{Discussion}

\chapter{Conclusion}

>>>>

Over the past sixty years, Artificial Neural Networks (ANNs) have experienced a number of popularity cycles that were noted by extensive publicity of over-inflated expectations with regards to the promises of connectionism (the field of science that tries to explain mental phenomena with the help of ANNs [CITATION NEEDED]) [CITATION NEEDED]. More specifically, it promised that a system could learn to do anything it was programmed to do \parencite{minsky_perceptrons_1988}. However, each wave of praise was followed by a mass of public disappointment, criticism, and funding cuts [CITATION NEEDED]. These upturns and downturns of the field of connectionism - or more generally A.I. - are now known in the literature as the A.I. winters [CITATION NEEDED].

The recent revival of ANNs under the scope of Deep Learning (DL) has brought new ways to tackle Machine Learning problems, most notably in the context of classification (but also regression, although a less popular application; see \cite{lecun_deep_2015} for several DL examples). In addition, with the advent of chatbots \parencite{dale_return_2016}, voice-activated personal assistants \parencite{xiong_microsoft_2018}, and general-purpose A.I. systems \parencite{vinyals_starcraft_2017}, DL introduced new frontiers in artificial intelligence research.

Broadly speaking, DL as a field, promises to solve tasks which are easily (intuitively) performed by people, but hard to formalize \parencite[e.g., recognizing faces or spoken word;][]{goodfellow_deep_2016}. The solution comprises of enabling computers to learn from experience and understand the world through the discovery of hierarchical relationships between concepts \parencite{lecun_deep_2015}. This way, there is no need for manual input on the sort of knowledge that the computer needs \parencite{goodfellow_deep_2016}. A more simplistic and concise interpretation of DL would be to see it as a statistical technique for identifying patterns in sample data using large (deep) ANNs \parencite{marcus_deep_2018}.

The reasons for presenting these facts will be clarified in the following sections (Section \ref{sec:thesis_goal} provides a concrete explanation) however, a terse argumentation is that progress towards general A.I. systems is marked by scant explorations of ANNs outside of the classification task, and even more so when trying to overcome some of their shortcomings, such as being highly reliant on massive amounts of data [CITATION NEEDED].
% NOTE: Maybe I was too harsh?

@@

For the purposes of this paper (detailed in Section \ref{sec:thesis_goal}), generative models are a central element in our experiments (Chapter \ref{chap:methods}). Borrowing the analogy from \textcite{goodfellow_nips_2016}, one might wonder what is the value of studying generative models. After all, in the image domain, such a framework would merely generate more images (something which the Internet as no shortage of). The answer to this lies in the fact that generative models can be used for simulating possible futures for scheduling and planning [CITATION NEEDED], providing predictions to missing data [CITATION NEEDED], or enabling work on multi-modal outputs, where a single input may have more than one correct answer \parencite{goodfellow_nips_2016}.

However, successes in generative models have been scarce, with Fully Visible Belief Networks \parencite{frey_does_1996} being the most suitable (and popular) method for the longest run. Yet, this method requires generating one sample at a time (ramping up the cost of the process to $O(n)$). Coupled with the fact that the computation is done through a neural network, creating $n$ samples requires a substantial amount of time \parencite{goodfellow_nips_2016}. Variational Autoencoders \parencite{kingma_auto-encoding_2013} are another popular method that is used for this sort of task. Their main drawback however, is that when a weak approximation of a prior or posterior distribution of the data is used, the model might learn something other than the true data distribution \parencite{goodfellow_nips_2016}.

@@

In addition, although not as popular as the abovementioned, Recurrent Neural Networks (RNNs) have been successfully implemented for various generative tasks \parencite{jenal_rnn-based_2019, ICML2011Sutskever_524}. More specifically, a variation of these, the Long Short-Term Memory Networks \parencite[LSTMs;][]{hochreiter_long_1997}, are of particular interest. RNNs (and by extension LSTMs) contain a high-dimensional hidden state which allows them to remember and predict from previous inputs \parencite{ICML2011Sutskever_524}. This makes them great candidates for predicting sequences \parencite{graves_generating_2013} of text or characters, where the input at $t_0$ might be influenced by prior input at $t_{-1}$, $t_{-2}$, etc. \parencite{fan_tts_2014}. That being said, the purpose-built memory cell of an LSTM permits better storing and accessing of information \parencite{hochreiter_long_1997, graves_generating_2013} which in turn makes this RNN variation a better candidate for this sort of task.

@@

Having said all of these, the majority of developments with generative models lie in the field of Computer Vision. But that is not to say that we can generate only images. Generative models have been applied with varying degrees of success also in Natural Language Processing - where the scope was to generate high-quality text with sufficient diversity \parencite[e.g.,][]{yu_seqgan_2016, chen_adversarial_2018}. Less explored, but still successful, have been the efforts in generating music \parencite[e.g.,][]{mogren_c-rnn-gan_2016, dong_musegan_2017}. Herein lies the scope of this paper. We evaluated the samples generated by a GAN \parencite{mogren_c-rnn-gan_2016} and LSTM \parencite{oore_this_2018} when trained under a meta-learning algorithm (see Section \ref{sec:hungry}). To reiterate, a full explanation of the purpose of the paper is provided in Section \ref{sec:thesis_goal}.

@@

DL models usually rely on a procedure called stochastic gradient descent (SGD) which implies computing an output and an error for a given input vector, followed by calculating the input's average gradient and adjusting the network's parameters in a manner that minimizes the error \parencite{lecun_deep_2015}. \textcite{ravi_optimization_2016} argue that the iterative nature of gradient-optimization algorithms does not allow them to perform well under the constraint of a set number of updates over few examples. In particular, when faced with non-convex optimization problems, DL algorithms do not guarantee convergence within a reasonable amount of time. In addition, they also emphasize that for each new problem (e.g., new dataset), the model needs to reinitialize with a new set of random parameters. Ultimately, this hurts the network's ability to converge to a good solution when constrained by a limited number of updates \parencite{ravi_optimization_2016}. Thus they lay out two clear drawbacks of DL models: "data-hunger", and inability to perform separate tasks.

A clear outcome from solving these issues would be that DL models are easier to train (less data) and multifunctional however, this list is not exhaustive. There is also fact that humans have the ability to generalize after one (or few) example(s) of a given object \parencite{vinyals_matching_2016, chen_closer_2018, ravi_optimization_2016} - something which DL models tend to lack. Moreover, there are many fields where the data exhibits a large number of classes, with few examples per class. Bridging the gap between human-type learning and current learning architectures would allow models to properly capture this sort of sparse data \parencite{ravi_optimization_2016, larochelle_few-shot_2017}. This has motivated a series of explorations in the direction of "few-shot" learning (i.e., learning a class from a few examples) or more broadly, into the field of "meta-learning" \parencite[learning to learn in hopes of generalizing to new tasks;][]{chen_closer_2018, vinyals_matching_2016, zhang_metagan_2018}.

In spite of its proposed benefits, developments and experiments with few-shot learning are still scarce \parencite{larochelle_few-shot_2017}. Moreover, evaluations are largely concentrated on image data \parencite[see][]{lake_omniglot_2019, clouatre_figr_2019, vinyals_matching_2016, chen_closer_2018, ravi_optimization_2016}. Having said that, few-shot experiments with generative networks are even less frequent; \textcite{clouatre_figr_2019}, \textcite{dong_musegan_2017} and \textcite{zhang_metagan_2018} being some of the early explorers in this direction.

\section{Composing like a human} \label{sec:thesis_goal}

At the time of writing, generative models and meta-learning are two exciting areas of research in DL which have been successfully combined. However, to the knowledge of the researcher, no such experiments have been conducted in the musical domain. Thus, this paper examined the extent to which generative models can create novel and qualitative samples of music under the constraints posed by few-shot learning. In other words, it sought to evaluate the extent to which generative models can create novel and high-fidelity samples when training data is scarce. To avoid ambiguity, we define novel as new (i.e., not in the training set), and qualitative as consistent with common scales and diverse (combines chord structures with single-note sequences, is not restricted to only a few notes, etc.). This research will evaluate the generated samples of two state-of-the-art generative models (C-RNN-GAN of \cite{mogren_c-rnn-gan_2016} and Performance-RNN of \cite{oore_this_2018}) that have been trained under a meta-learning framework, Reptile \parencite{nichol_first-order_2018}, on a limited number of piano recordings. The performances of these two models were compared to the baseline proposed by \textcite{larochelle_few-shot_2017} and to recorded performances played by expert piano players at the International e-Piano Competition \parencite{university_of_minnesota_international_2019}.

In other words, this study sought to answer the following research questions:

\begin{itemize}
    \item To what extent is the music created by a few-shot generative model comparable to the music of a generative model that is trained on the entire dataset?
    \item To what extent is the music created by a few-shot generative model comparable to real music?
\end{itemize}

The following chapters will discuss more about generative models, meta-learning, and some of the models and algorithms involved in this paper (Chapter \ref{chap:theoretical_framework}), followed by a detailed description of the setup of the experiments (Chapter \ref{chap:methods}), and the results obtained from them (Chapter \ref{chap:results}). Finally, Chapter \ref{chap:discussion} closes with a discussion of the results and limitations of the experimental setup, whereas Chapter \ref{chap:conclusion} draws some conclusions that answer the above research questions.

\chapter{Related Work}\label{chap:theoretical_framework}

@@

@@

\chapter{Results}\label{chap:results}

\chapter{Discussion}\label{chap:discussion}

\chapter{Conclusion}\label{chap:conclusion}
